---
title: "IntroR6"
---

# Central Limit Theorem

Consider a sample of random coin tosses. Each coin toss is sampled from a Bernouli distribution with p=.5, or equivalently from a Binomial distribution of size 1 and p=0.5. The mean of the Bernoulli $\mu$ is 0.5 and the standard deviation $\sigma$ is 0.5 (see Wikipedia). The central limit theorem says that the mean value of the sample of size n has a normal distribution with mean equal to μ and standard deviation equals to $\sigma /\sqrt(n)$.

```{r}

nmax=10000
nsamples=100
m=0
n=0
p = 0.05
mean = p
stdev = sqrt(p*(1-p))
for (i in 1:nmax) # number of means calculated
  {
  x = rbinom(nsamples, 1, .05) # number of values draw from distribution
  m[i] = mean(x)  # mean of the sample
  n[i] = (mean(x)-mean)/(stdev/sqrt(nsamples))  # normalized sample mean
}
hist(m)
mean(m)
sd(m)
.5/sqrt(nsamples)
hist(n)
sort(m)[0.025*nmax] #lower bound confidence interval 95%
sort(m)[0.975*nmax] #upper bound confidence interval 95%
qnorm(0.025,mean=.5,sd=0.5/sqrt(nsamples))
qnorm(0.975,mean=.5,sd=0.5/sqrt(nsamples))
mean(n)
sd(n)
sort(n)[0.025*nmax] #lower bound confidence interval 95%
sort(n)[0.975*nmax] #upper bound confidence interval 95%
```

```{r}
qnorm(0.025,mean=0,sd=1)
qnorm(0.975,mean=0,sd=1)
```
