## Ecophysiology of an ectotherm

Solar radiation and convection are the two main pathways for animals such as small lizards. In this case, the total heat flux (*f*) into the animal is given by the **heat flux equation**:

$$f = q\ –h*(b - a)$$

-   $q$ = solar radiation (cal/h)

-   $h*(b - a)$ = heat loss through convection

-   $b$ = body temperature of the animal (ºC)

-   $a$ = air temperature (ºC)

-   $h$ = convection heat transfer coefficient (cal/h/ ºC)

<!--# I have modified this to change air temperature and h so that's exactly not the same as in the book -->

1.  ***Plotting the cimate space***

    a.  Assuming that air temperature is 21ºC, and **h** = 60 cal/h, plot the lizard's equilibrium body temperature as a function of solar radiation. Assume that solar radiation varies between 0 and 1500 cal/h

    b.  Draw the climate space of the lizard by drawing the polygon that is limited by the minimum and maximum temperatures that the lizard can support, as a function of solar radiation. Consider that the upper lethal limit for the body temperature (***bmax***) is 36ºC and the lower lethal limit for the body temperature (***bmin***) is 24ºC.

    c.  Plot the following values of air temperature and solar radiation that are available throughout a day in two locations.

    ```{r eval=FALSE}
    t <- c("00:00","03:00","06:00","09:00",    # Times of the day 
       "12:00","15:00","18:00","21:00")

    # Rock habitat
    rock_q <- c(150,150,800,1100,1300,1200,800,400,150) # solar radiation
    rock_a <- c(18,13,10,14,21,24,22,20,18)             # air temperature

    # Bush habitat
    bush_q <- c(150,150,450,600,650,650,350,200,150)    # solar radiation
    bush_a <- c(18,13,10,14,21,24,22,20,18)             # air temperature
    ```

    d.  At which time of the day is the lizard on the rock, and at which time is it at the bush? Is there any time when the lizard cannot be at any of the locations?

2.  ***Statistcal modelling of body temperature***

@sec-anolis_cristatellus presents a data frame containing data on individuals *A. cristatellus* living in urban and forested areas in three cities in Puerto Rico, giving us the perfect opportunity to assess the potential thermal effects of habitat upon this species. To do this:

a.  Make a linear model for individuals of *A. cristatellus* living only in natural habitats (`context` = `natural`) and another for individuals living in urban habitats (`context` = `urban`).

**Hint:** to filter the data you can use the `subset` function in R

```{r eval=FALSE}
# example filtering data from anoles only from San Jose sites
df_sanjose <- subset(df, Site == "San Jose")
```

b.  Is there a habitat where lizards are consistently warmer? If so, provide an ecological explanation (quantitatively) for this observation (assume an ambient temperature of 30 °C).

c.  Is your model appropriate for explaining your data?
